Secure Composition of Robust and Optimising Compilers

### 2024-10-11 tech/main.tex
tech

@@

\gls*{tms}~\cite{nagarakatte2010cets} is a safety property that describes that an unallocated object must not be (re-)used.

\begin{definition}{\definitionlabel[\glsfirst{tms}]{def-trace-tmsdef}}
  \[
  \tmssafe:=\left\{\varTrace[_{\mssafe}]
    \left| 
      \begin{array}{rcl}
        \ev{Alloc\ \varLoc\ n}&\le_{\varTrace[_{\mssafe}]}&\ev{Dealloc\ \varLoc} \\
        \ev{Use\ \varLoc\ n}&\le_{\varTrace[_{\mssafe}]}&\ev{Dealloc\ \varLoc} \\
        \text{at most one }\ev{Dealloc\ \varLoc}&\text{in}&\varTrace[_{\mssafe}] \\
        \text{at most one }\ev{Alloc\ \varLoc\ n}&\text{in}&\varTrace[_{\mssafe}] \\
      \end{array}
    \right.
  \right\}
  \]
\end{definition}
Hereby, the notation $\varEvent[_{1}]\le_{\varTrace}\varEvent[_{2}]$ means that if $\varEvent[_{1}]$ is in $\varTrace$ and if $\varEvent[_{2}]$ is in $\varTrace$, then $\varEvent[_{1}]$ appears before $\varEvent[_{2}]$.

\paragraph{Spatial Memory Safety}
\gls*{sms}~\cite{nagarakatte2009soft} is a safety property that prohibits out-of-bounds accesses.

\begin{definition}{\definitionlabel[\glsfirst{sms}]{def-trace-smsdef}}
  \[
  \smssafe:=\left\{\varTrace[_{\mssafe}] \left|\begin{array}{rcl}
      \text{If }\ev{Alloc\ \varLoc\ n}\le_{\varTrace_{\mssafe}}\ev{Use\ \varLoc\ m}, \text{ then }m<n
  \end{array}\right.\right\}
  \]
\end{definition}

@@

In spirit of earlier work~\cite{nagarakatte2009soft,nagarakatte2010cets,jim2002cyclone,necula2005ccured,michael2023mswasm}, full \gls*{ms} is the intersection of \Cref{def:trace:tmsdef,def:trace:smsdef}.

\begin{definition}{\definitionlabel[\glsfirst{ms}]{def-trace-msdef}}
  \[
  \mssafe:=\tmssafe \cap \smssafe
  \]
\end{definition}

Note that \Cref{def:trace:msdef} ignores data isolation, so there may still be memory-safety issues introduced by side-channels.

@@

The combination of \gls*{ms} and \gls*{scct} is the intersection of these properties, \gls*{msscct}.
However, \gls*{ms} uses a different trace model than \gls*{scct}, so intersecting them would trivially yield the empty set. 
To remedy this issue, we introduce $\sim_{\ctsafe}: \preevent[_{\ctsafe}] \times \preevent[_{\mssafe}] $, a cross-language trace relation (whose key cases are presented below), that we use to intuitively unify the trace model in which the two properties are expressed:
\[
  \typerulenolabel{}{\varEvent[_{\mssafe}]\sim_{\ctsafe}\varEvent[_{\mssafe}];\unlock{}}
  \hspace{2ex}
  \typerulenolabel{}{\lightning\sim_{\ctsafe}\lightning}
  \hspace{2ex}
  \typerulenolabel{}{\emptyevent\sim_{\ctsafe}\ev{Branch\;n};\varSecuritytag{}}
  \hspace{2ex}
  \typerulenolabel{}{\emptyevent\sim_{\ctsafe}\ev{Binop\;n};\varSecuritytag{}}
\]

@@

\begin{definition}{\definitionlabel[\glsfirst{ss}]{def-trace-ssdef}}
  \noindent

  \begin{nscenter}
    \[
      \sssafe := \left\{
        \varTrace[_{\mathghost}] 
          \left|
            \begin{array}{l}
              \varTrace[_{\mathghost}]=\hole{\cdot} 
                \text{ or } 
              \exists\varTrace[_{\mathghost}'].\\
              %
              \left(\varTrace[_{\mathghost}]=\varEvent[_{\mathghost}];\unlock\cdot\varTrace[_{\mathghost}'] 
                \text{or }
              \varTrace[_{\mathghost}]=\varEvent[_{\mathghost}];\lock_{\text{NONE}}\cdot\varTrace[_{\mathghost}']\right)\\
              %
              \text{and }%
              \ \varTrace[_{\mathghost}']\in\sssafe
            \end{array}
          \right.
        \right\}
    \] 
  \end{nscenter}
\end{definition}
The technical setup so far leads to the above definition, where only locks annotated with $\text{SPECTRE-PHT}$ are disallowed to occur on the trace.
That way, programs attaining \gls*{ss} do not necessarily attain \gls*{scct}.

>>>>

For simple memory safety composed of temporal and spatial memory safety, the trace model defines events ($\varEvent[_{\mssafe}]$) as either the empty event ($\emptyevent$), a crash ($\lightning$), or a base-event ($\varEvent[_{\mssafe}]$).

\begin{definition}{\definitionlabel[\glsfirst{ms} Trace Model]{mstraces}}
\[
  \begin{array}{rcll}
    \text{(Base-Events)}&\varEvent[_{\mssafe}] &:=& \msAlloc \mid \msDealloc \mid \msUse \\
    \text{(Events)}&\varEvent[_{\mssafe}] &:=& \varEvent[_{\mssafe}] \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]
\end{definition}

Base-events describe the actual kind of event that happened.
For the basic memory-safety properties, these are three variants:
First, the allocation event ($\msAlloc$) that fires whenever a program claims $n$ cells of memory and stores them at address $\varLoc$, where addresses are assumed to be unique.
Second, deallocation ($\msDealloc$) announces that the object at location $\varLoc$ is freed.
Third, an event to describe reads from and writes to the $n$-th memory cell from address $\varLoc$ ($\msUse$).

@@

\gls*{tms}~\cite{nagarakatte2010cets} is a safety property that describes that an unallocated object must not be (re-)used.
Since it does not care about bounds of an object in memory, we emphasize this fact by deviating from the unified, memory-safety trace model and instead introduce traces for temporal memory safety:
\begin{definition}{\definitionlabel[\glsfirst{tms} Traces]{tmstraces}}
\[
  \begin{array}{rcll}
    \text{(Base-Events)}&\tmsEvent &:=& \tmsAlloc \mid \tmsDealloc \mid \tmsUse \\
    \text{(Events)}&\tmsEvent &:=& \tmsEvent \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]
\end{definition}

With this, \gls*{tms} is defined as follows.
\begin{definition}{\definitionlabel[\glsfirst{tms}]{tmsdef}}
  \[
  \tmssafe:=\left\{\tmsTrace
    \left| 
      \begin{array}{rcl}
        \tmsAlloc&\le_{\tmsTrace}&\tmsDealloc \\
        \tmsUse&\le_{\tmsTrace}&\tmsDealloc \\
      \end{array}
    \right.
  \right\}
  \]
\end{definition}
Hereby, the notation $\varEvent[_{1}]\le_{\varTrace}\varEvent[_{2}]$ means that if $\varEvent[_{1}]$ is in $\varTrace$ and if $\varEvent[_{2}]$ is in $\varTrace$, then $\varEvent[_{1}]$ appears before $\varEvent[_{2}]$.

\paragraph{Spatial Memory Safety}
\gls*{sms}~\cite{nagarakatte2009soft} is a safety property that prohibits out-of-bounds accesses.
For emphasis, we also use a custom trace-model for spatial memory safety.
\begin{definition}{\definitionlabel[\glsfirst{sms} Traces]{smstraces}}
\[
  \begin{array}{rcll}
    \text{(Base-Events)}&\smsEvent &:=& \smsAlloc \mid \smsDealloc \mid \smsUse \\
    \text{(Events)}&\smsEvent &:=& \smsEvent \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]
\end{definition}

With this, \gls*{sms} is defined as follows.
\begin{definition}{\definitionlabel[\glsfirst{sms}]{smsdef}}
  \[
  \smssafe:=\left\{\smsTrace \left|\begin{array}{rcl}
      \text{If }\smsAlloc\le_{\smsTrace}\smsUse, \text{ then }m<n
  \end{array}\right.\right\}
  \]
\end{definition}

@@

In spirit of earlier work~\cite{nagarakatte2009soft,nagarakatte2010cets,jim2002cyclone,necula2005ccured,michael2023mswasm}, full \gls*{ms} is the intersection of \definitionref{tmsdef} and \definitionref{smsdef}.
However, both properties use a different trace model, so intersecting them alone would trivially yield the empty set.
To remedy this, we define cross-language trace relations that lift both \definitionref{tmstraces} and \definitionref{smstraces} to \definitionref{mstraces}.
\begin{definition}{\definitionlabel[Mapping \glsfirst{ms} to \glsfirst{tms}]{mapping-tms-ms}}
  \newcommand{\mmap}[2]{{#1}\;{\xlangrel{\mssafe}{\tmssafe}}\;{#2}}
  \begin{center}
    \judgbox{\mmap{\msEvent}{\tmsEvent}}{,,Translate \gls*{ms}-events to \gls*{tms}-events.''}

    \typerule{mapping-ms-tms-empty}{}{
      \mmap{\msev{\emptyevent}}{\tmsev{\emptyevent}}
    }{mapping-ms-tms-empty}
    %
    \typerule{mapping-ms-tms-crash}{}{
      \mmap{\msev{\lightning}}{\tmsev{\lightning}}
    }{mapping-ms-tms-crash}
    %
    \typerule{mapping-ms-tms-alloc}{}{
      \mmap{\msAlloc}{\tmsAlloc}
    }{mapping-ms-tms-alloc}
    %
    \typerule{mapping-ms-tms-dealloc}{}{
      \mmap{\msDealloc}{\tmsDealloc}
    }{mapping-ms-tms-dealloc}
    %
    \typerule{mapping-ms-tms-use}{}{
      \mmap{\msUse}{\tmsUse}
    }{mapping-ms-tms-use}
  \end{center}
  \begin{center}
    \judgbox{\mmap{\msTrace}{\tmsTrace}}{,,Translate \gls*{ms}-traces to \gls*{tms}-traces.''}

    \typerule{mapping-ms-tms-empty}{}{
      \mmap{\emptyTrace}{\emptyTrace}
    }{mapping-ms-tms-empty}
    %
    \typerule{mapping-ms-tms-cons}{}{
      \mmap{\consTrace{\msEvent}{\msTrace}}{\consTrace{\tmsEvent}{\tmsTrace}}
    }{mapping-ms-tms-cons}
  \end{center}
\end{definition}
\begin{definition}{\definitionlabel[Mapping \glsfirst{ms} to \glsfirst{sms}]{mapping-sms-ms}}
  \newcommand{\mmap}[2]{{#1}\;{\xlangrel{\mssafe}{\smssafe}}\;{#2}}
  \begin{center}
    \judgbox{\mmap{\msEvent}{\smsEvent}}{,,Translate \gls*{ms}-events to \gls*{sms}-events.''}

    \typerule{mapping-ms-sms-empty}{}{
      \mmap{\msev{\emptyevent}}{\smsev{\emptyevent}}
    }{mapping-ms-sms-empty}
    %
    \typerule{mapping-ms-sms-crash}{}{
      \mmap{\msev{\lightning}}{\smsev{\lightning}}
    }{mapping-ms-sms-crash}
    %
    \typerule{mapping-ms-sms-alloc}{}{
      \mmap{\msAlloc}{\smsAlloc}
    }{mapping-ms-sms-alloc}
    %
    \typerule{mapping-ms-sms-dealloc}{}{
      \mmap{\msDealloc}{\smsDealloc}
    }{mapping-ms-sms-dealloc}
    %
    \typerule{mapping-ms-sms-use}{}{
      \mmap{\msUse}{\smsUse}
    }{mapping-ms-sms-use}
  \end{center}
  \begin{center}
    \judgbox{\mmap{\msTrace}{\smsTrace}}{,,Translate \gls*{ms}-traces to \gls*{sms}-traces.''}

    \typerule{mapping-ms-sms-empty}{}{
      \mmap{\emptyTrace}{\emptyTrace}
    }{mapping-ms-sms-empty}
    %
    \typerule{mapping-ms-sms-cons}{}{
      \mmap{\consTrace{\msEvent}{\msTrace}}{\consTrace{\smsEvent}{\smsTrace}}
    }{mapping-ms-sms-cons}
  \end{center}
\end{definition}

\begin{definition}{\definitionlabel[\glsfirst{ms}]{msdef}}
  \[
    \mssafe:=\mapUniversal{\xlangrel{\mssafe}{\tmssafe}}{\tmssafe} \cap \mapUniversal{\xlangrel{\mssafe}{\smssafe}}{\smssafe}
  \]
\end{definition}

Note that \definitionref{msdef} ignores data isolation, so there may still be memory-safety issues introduced by side-channels.

@@

The combination of \gls*{ms} and \gls*{scct} is the intersection of these properties, \gls*{msscct}.
However, \gls*{ms} uses a different trace model than \gls*{scct}, so intersecting them would trivially yield the empty set. 
To remedy this issue, we introduce $\sim_{\ctsafe}: \varEvent[_{\ctsafe}] \times \varEvent[_{\mssafe}] $, a cross-language trace relation (whose key cases are presented below), that we use to intuitively unify the trace model in which the two properties are expressed:
\[
  \typerulenolabel{}{\varEvent[_{\mssafe}]\sim_{\ctsafe}\varEvent[_{\mssafe}];\unlock{}}
  \hspace{2ex}
  \typerulenolabel{}{\lightning\sim_{\ctsafe}\lightning}
  \hspace{2ex}
  \typerulenolabel{}{\emptyevent\sim_{\ctsafe}\ev{Branch\;n};\varSecuritytag{}}
  \hspace{2ex}
  \typerulenolabel{}{\emptyevent\sim_{\ctsafe}\ev{Binop\;n};\varSecuritytag{}}
\]

@@

\begin{definition}{\definitionlabel[\glsfirst{ss}]{def-trace-ssdef}}
  \[
    \sssafe := \left\{
      \varTrace[_{\mathghost}] 
        \left|
          \begin{array}{l}
            \varTrace[_{\mathghost}]=\hole{\cdot} 
              \text{ or } 
            \exists\varTrace[_{\mathghost}'].\\
            %
            \left(\varTrace[_{\mathghost}]=\varEvent[_{\mathghost}];\unlock\cdot\varTrace[_{\mathghost}'] 
              \text{or }
            \varTrace[_{\mathghost}]=\varEvent[_{\mathghost}];\lock_{\text{NONE}}\cdot\varTrace[_{\mathghost}']\right)\\
            %
            \text{and }%
            \ \varTrace[_{\mathghost}']\in\sssafe
          \end{array}
        \right.
      \right\}
  \] 
\end{definition}
The technical setup so far leads to the above definition, where only locks annotated with $\text{SPECTRE-PHT}$ are disallowed to occur on the trace.
That way, programs attaining \gls*{ss} do not necessarily attain \gls*{scct}.
